---
title: "cci-getting-started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{cci-getting-started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(CCI)
```
# Overview

The CCI package provides a computational test for conditional independence (CI) that leverages machine learning and Monte Carlo resampling.
We test statements of the form 𝑌⊥ X ∣𝑍 by comparing predictive performance under permutation to build an empirical (or parametric) null.

This vignette shows:
- how to run a basic CI test,
- how CCI adapts to outcome type (continuous vs. categorical),
- how to interpret output and visualize the null distribution,
- a few practical knobs: metric, nperm, p, choose_direction.

Throught these examples, we use simulated data where the ground truth is known.
## Basic test
The function below generates a data set where `Y` is conditionally independent of `X` given `Z1` and `Z2`. The functions are linear 
so it should be easy to test conditional independence using traditional statistical tests as well.
```{r}
NormalData <- function(N){
  Z1 <- stats::rnorm(N,0,1)
  Z2 <- stats::rnorm(N,0,1)
  X <- stats::rnorm(N, Z1 + Z2, 1)
  Y <- stats::rnorm(N, Z1 + Z2, 1)

  df <- data.frame(Z1, Z2, X, Y)
  return(df)
}
```
The simplest line of code needed to test the hypothesis `Y _||_ X | Z1, Z2` is: 
```{r}
set.seed(123)
dat <- NormalData(500)
summary(CCI.test(Y ~ X | Z1 + Z2, data = dat))
```
This runs a conditional independence test using the default settings: random forests as the machine learning method, RMSE as the metric. 
The default settings is quit robust and works well in many scenarios, but the user can customize them as needed (examples below).
```{r}
HardCase <- function(N) {
  Z1 <- stats::runif(N, -2, 2)
  Z2 <- stats::runif(N, -2, 2)
  hZ <- sin(Z1) * cos(Z2)
  X <- hZ + 0.2 * stats::rnorm(N)
  Y <- hZ^2 + 0.2 * stats::rnorm(N)
  data.frame(X, Y, Z1, Z2)
}
```

```{r}
set.seed(1)
dat <- HardCase(500)
summary(CCI.test(Y ~ X | Z1 + Z2, data = dat))
summary(CCI.test(Y ~ X | Z1, data = dat)) # Shouid be rejected
```

## Adapting to outcome type
CCI automatically adapts to the outcome type. For continuous outcomes, it uses regression methods and metrics RMSE; for categorical outcomes, it uses classification methods and metrics Kappa.
```{r}
ExpLogThreshold <- function(N) {
  Z1 <- stats::rnorm(N)
  Z2 <- stats::rnorm(N)
  X <- exp(Z1) + Z2 + stats::rnorm(N, 0, 0.2)
  Y <- ifelse(log(abs(Z1) + 1) + Z2 > 0.5, "Goblin",
              ifelse(log(abs(Z1) + 1) + Z2 > 0, "Orc",
                     ifelse(log(abs(Z1) + 1) > -0.5, "Troll", "Elf")))
  return(data.frame(Z1, Z2, X, Y))
}
```
By setting the outcome `Y` as a factor, CCI recognizes it as a classification problem and automatically switches to classification methods and metrics.
```{r}
set.seed(123)
dat <- ExpLogThreshold(500)
dat$Y <- as.factor(dat$Y)
summary(CCI.test(Y ~ X | Z1 + Z2, data = dat))
summary(CCI.test(Y ~ X | Z1, data = dat)) 
```
The user can also manually specify the metric to use via the `metric` argument.
```{r}
set.seed(123)
dat <- ExpLogThreshold(500)
dat$Y <- as.numeric(as.factor(dat$Y))
summary(CCI.test(Y ~ X | Z1 + Z2, data = dat, metric = "Kappa"))
```
## Visualizing the null distribution
The null distribution can be visualized using the `plot` function.
```{r}
set.seed(13)
dat <- NormalData(500)
cci <- CCI.test(Y ~ X | Z1 + Z2, data = dat)
summary(cci)
plot(cci)
```
The dashed line indicates the observed test statistic, while the histogram represents the null distribution generated through permutations.
## Practical knobs
### Choosing method
The user can choose the machine learning method via the `method` argument. Supported methods include "rf" (random forests), "xgboost", and support vector machines "svm". Users should also use the `seed` argument to ensure reproducibility when using stochastic methods.
```{r}
summary(CCI.test(Y ~ X | Z1, data = dat, method = "xgboost", seed = 99))
summary(CCI.test(Y ~ X | Z1, data = dat, method = "svm", seed = 99))
```
A third argument which beneficial should be set to TRUE is the parametric argument. This will compute a parametric p-value based on the null distribution being approximately Gaussian. 
```{r}
summary(CCI.test(Y ~ X | Z1, data = dat, method = "xgboost", seed = 99, parametric = TRUE))
```



